---
title: "Tests / Examples using glmmTMB with DHARMa"
author: "Florian Hartig"
date: "3/27/2018"
output: 
  html_document: 
    keep_md: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = T)
```

```{r}
library(DHARMa)
library(glmmTMB)
```

# Testing general package behavior / general problems


## Pearson residuals don't work with zi terms (not implemented)

## with ~reform in pred / simulate

Currently, glmmTMB doesn't support the reform argument. Also, 

* predict() is conditional on all random effects, corresponding to lme4 re.form = NULL
* simulate() is unconditional, i.e. all random effects will be resimulated, corresponding to lme4 re.form = 0

all predictions and simulations are conditional on REs, problematic for some of the residual tests, see https://github.com/florianhartig/DHARMa/issues/43

```{r}
m1 <- glmmTMB(count~ mined + (1|site), family=poisson, data=Salamanders)


#pred = predict(m1, re.form = 0)
boxplot(Salamanders$mined, predict(m1, type = c("link")))
boxplot(Salamanders$mined, predict(m1, type = c("response")))

```


# DHARMa with glmmTMB example cases 

## Test 1 - Simple poisson model

```{r}
m <- glmmTMB(count~ mined, family=poisson, data=Salamanders)
summary(m)

res = simulateResiduals(m)
plot(res, asFactor = T)

### creating new data based on the fitted models. Residuals should now be perfect

Salamanders$count2 = simulate(m)$sim_1

m <- glmmTMB(count2~ mined, family=poisson, data=Salamanders)

res = simulateResiduals(m)
plot(res, asFactor = T)
```


## Test 2 - Poisson + RE

```{r}
m <- glmmTMB(count~ mined + (1|site), family=poisson, data=Salamanders)
summary(m)

res = simulateResiduals(m)
plot(res, rank = T)

### creating new data based on the fitted models. Residuals should now be perfect, except for the reform pattern

Salamanders$count2 = simulate(m)$sim_1

m <- glmmTMB(count2~ mined + (1|site), family=poisson, data=Salamanders)

res = simulateResiduals(m)
plot(res, rank = T)
```

Fit is still fine, but note the pattern in the residual vs. predicted, which is caused by the reform problem (REs are included in the predictions)

If we calculate unconditional REs by hand, pattern disappears

```{r}
x = fixef(m)
pred = x$cond[1] + x$cond[2]*as.numeric(Salamanders$mined)
plotResiduals(pred, res$scaledResiduals, asFactor = T)
```



## Test 3 - Binomial models

Binomial required a few adjustments, because the simulate function in glmmTMB doesn't behave like glm / glmer, which returns lists of for the n/k case, and a vector instead

standard binomial with DHARMa test data

```{r}
testData = createData(sampleSize = 100, fixedEffects = 2, family = binomial(), randomEffectVariance = 0)

m <- glmmTMB(observedResponse ~ Environment1 , data=testData, family=binomial)
res = simulateResiduals(m)
plot(res)
```

with a missing predictor

```{r}
testData = createData(sampleSize = 100, fixedEffects = 2, family = binomial(), randomEffectVariance = 0)

m <- glmmTMB(observedResponse ~ 1 , data=testData, family=binomial)
res = simulateResiduals(m)
plot(res)
plotResiduals(testData$Environment1, res$scaledResiduals)
```

n/k example from the help

```{r}
data(cbpp, package="lme4")

newDat = cbpp
newDat$response = cbind(newDat$incidence, newDat$size-newDat$incidence)
m <- glmmTMB(response ~ period , data=newDat, family=binomial)
m

res = simulateResiduals(m)
plot(res, rank = T)

### creating new data based on the fitted models. Residuals should now be perfect

# R problem - sometimes newDat$newResponse is updated, and sometimes not ... I have no idea why

newDat$newResponse = as.matrix(simulate(m))
newDat$newResponse
m2 <- glmmTMB(newResponse ~ period , data=newDat, family=binomial)
m2

res = simulateResiduals(m2)
plot(res, rank = T)
```


## Test 4 - Zero-inflated negative binomial models

no RE to test ZI intercept only

```{r}
m <- glmmTMB(count~1 , zi=~1, family=nbinom1, Salamanders)
summary(m)

res = simulateResiduals(m)
plot(res, asFactor = T)

### creating new data based on the fitted models. Residuals should now be perfect

Salamanders$count2 = simulate(m)$sim_1
m <- glmmTMB(count2~1 , zi=~1, family=nbinom1, Salamanders)
summary(m)

res = simulateResiduals(m)
plot(res, asFactor = T)
```

Example from the help

```{r}
m <- glmmTMB(count~spp + mined + (1|site), zi=~spp + mined, 
             family=nbinom2, Salamanders)
summary(m)

res = simulateResiduals(m)
plot(res)

### creating new data based on the fitted models. Residuals should now be perfect

Salamanders$count2 = simulate(m)$sim_1
m <- glmmTMB(count~spp + mined + (1|site), zi=~spp + mined, family=nbinom2, Salamanders)
summary(m)

res = simulateResiduals(m)
plot(res, rank = T)
```

There is still a slight correlation in both cases - I assume that this is again a problem of https://github.com/florianhartig/DHARMa/issues/43

# Model types

## Tweedie

Tweedie distributions are continous, but can have point mass at zero, which makes them effectively semi-discrete - this can create patterns in the residuals if one simulates as for a continous distribution, see [here](https://github.com/florianhartig/DHARMa/issues/66).

DHARMa therefore checks if there are repeated values in the simulations / observed data, and if so, switches to an integer distribution

```{r}
# Simulate Tweedie data
set.seed(16)
dat = data.frame( y = tweedie::rtweedie(200, mu = 2, phi = 2, p = 1.5),
                  x = rnorm(200) )

# Fit a model with single, unrelated explanatory variable
fit1 = glmmTMB(y ~ x, data = dat,
               family = tweedie)

summary(fit1)

dat = data.frame( y = simulate(fit1, 200)$sim_1,
                  x = dat$x )

fit1 = glmmTMB(y ~ x, data = dat,
               family = tweedie)
simres = simulateResiduals(fittedModel = fit1)

plot(simres)


```


## Beta-binomial

```{r}

testData = createData(sampleSize = 200, overdispersion = 0, randomEffectVariance = 0, family = binomial(), binomialTrials = 20)
            

fittedModel <- glmmTMB(cbind(observedResponse1,observedResponse0) ~ Environment1 + (1|group) , family = "binomial", data = testData)


fittedModel <- glmmTMB(cbind(observedResponse1,observedResponse0) ~ Environment1 + (1|group) , family = "betabinomial", data = testData)

res = simulateResiduals(fittedModel)
plot(res, rank = F)




```



## Beta

Beta distribution seems to work fine

```{r}
dat = data.frame(x = runif(200), y = rbeta(200, 0.5, 0.5))
m1 <- glmmTMB(y~ x , family=beta_family, data=dat)
summary(m1)

#m1$modelInfo$family$variance(1,2)
simulate(m1)

res <- simulateResiduals(m1)
plot(res)

```


# Old problems

## Problems with binomial

This used to create crashes my R (memory allocation issue) on the current CRAN version (27.3.18). Installing the development version on GitHub fixes the problem. sessionInfo() at the end of the document

-> update ... seeems to be fine now

```{r, eval = F}
data(cbpp, package="lme4")
m <- glmmTMB(cbind(incidence, size-incidence) ~ period + (1 | herd), data=cbpp, family=binomial)
summary(m)
predict(m)
```

## n/k update problem 

update doesn't work for n/k data

```{r, eval = F}
testData = createData(sampleSize = 200, family = binomial(), binomialTrials = 20)

fittedModel <- glmmTMB(cbind(observedResponse1,observedResponse0) ~ Environment1 + (1|group) , family = "binomial", data = testData)
temp = model.frame(fittedModel)
update(fittedModel, temp)
```

soluation -> DHARMa has now a refit function implemented for glmmTMB

## residual function doesn't work with factor response (bug)

```{r, eval = F}
testData = createData(sampleSize = 200, family = binomial(), factorResponse = T)
fittedModel <- glmmTMB(observedResponse ~ Environment1 + (1|group) , family = "binomial", data = testData)
residuals(fittedModel)

```

solved



# Reproducibility info 

```{r}
sessionInfo()
```


Hi guys, I'm testing glmmTMB for DHARMA, and I'm encountering a problem with the development versin on GitHub, the following code runs with the CRAN version, but not with the GitHub version

https://github.com/glmmTMB/glmmTMB/commit/b0084337f92736dfe84fd0f6386c1180a6103181


```{r}
library(DHARMa)
library(glmmTMB)
testData = createData(sampleSize = 500, intercept = 2, fixedEffects = c(1), overdispersion = 0, family = poisson(), quadraticFixedEffects = c(-3), randomEffectVariance = 0, pZeroInflation = 0.6)
fittedModel <- glmmTMB(observedResponse ~ Environment1 + I(Environment1^2) + (1|group), ziformula = ~1 , family = "poisson", data = testData)
summary(fittedModel)
predict(fittedModel, type = "response") 

simulationOutput <- simulateResiduals(fittedModel = fittedModel)
plot(simulationOutput)
```




