---
title: "Beetles example"
author: "Florian Hartig"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
abstract: "This example demonstrates how to create and read in simulated residuals from a Bayesian analysis in DHARMa"

---


# Create data

This first part creates a dataset with beetles counts across an altitudinal gradient (several plots each observed several years), with a random intercept on year and zero-inflation. 

```{r}
altitude = rep(seq(0,1,len = 50), each = 20)
dataID = 1:1000
spatialCoordinate = rep(seq(0,30, len = 50), each = 20)

# random effects + zeroinflation
plot = rep(1:50, each = 20)
year = rep(1:20, times = 50)

yearRandom = rnorm(20, 0, 1)
plotRandom = rnorm(50, 0, 1)
overdispersion = rnorm(1000, sd = 0.5)
zeroinflation = rbinom(1000,1,0.6)

beetles <- rpois(1000, exp( 0  + 12*altitude - 12*altitude^2 
                            #  + overdispersion   + plotRandom[plot]
                            + yearRandom[year]) * zeroinflation )

data = data.frame(dataID, beetles, altitude, plot, year, spatialCoordinate)

plot(year, altitude, cex = beetles/50, pch =2, main = "Beetle counts across altitudinal gradient\n triangle is proportional to counts")
```

# Analysis and residual analysis with JAGS

Prepare analysis

```{r}
library(R2jags)
modelData=as.list(data)
modelData = append(data, list(nobs=1000, nplots = 50, nyears = 20))
head(data)

```

## Fit GLM only 

```{r}

modelstring="
model {

# Likelihood
for (i in 1:nobs) {
  lambda[i] <- exp(intercept + alt * altitude[i] + alt2 * altitude[i] * altitude[i]) 
  beetles[i]~dpois(lambda[i]) 
}

# Fixed effect priors 
intercept ~ dnorm(0,0.0001)
alt ~ dnorm(0,0.0001)
alt2 ~ dnorm(0,0.0001)

# Posterior predictive simulations 

for (i in 1:nobs) {
  beetlesPred[i]~dpois(lambda[i])
}
Prediction <- sum(beetlesPred)
}
"

model=jags(model.file = textConnection(modelstring), data=modelData, n.iter=10000,  parameters.to.save = c("intercept", "alt", "alt2", "beetlesPred", "lambda"), DIC = F)

library(DHARMa)
simulations = model$BUGSoutput$sims.list$beetlesPred
pred = apply(model$BUGSoutput$sims.list$lambda, 2, median)
dim(simulations)
sim = createDHARMa(simulatedResponse = t(simulations), observedResponse = data$beetles, fittedPredictedResponse = pred, integerResponse = T)
plotSimulatedResiduals(sim)

```


We clearly see overdispersion in the plots.


# GLMM with random intercept on year, observation-level RE for overdispersion, and zero-inflation

The following code fits a GLMM with random intercept on year, observation-level RE for overdispersion, and zero-inflation.

Note that we then created two types of posterior predictive simulations 

* Conditional posterior predictive simulations, meaning that we take the final fitted values for lampda (which includes destimates of REs and ZI), and only simulate the final Poisson distribution

* Unconditional simulations, which means that we created data from the fitted model, simulating also the REs and the ZI terms newly

Below, I demonstrate both. If we want to create fully uncoditional simulations, we can either copy the entire model structure (as done in this example), or (small trick), we just make a copy of the data, merge it with the existing data, and set the response to NA - this way Jags will see sample the empty notes according to the same structure that is used to fit the data. 

Of course, with more complicated models, one can also do a part of the structure conditional, and another part unconditional. For example, here, we could have made the observation-level RE unconditional (as this is essentially a part of the distribution), but kept the REs on year and the zero-inflation conditional. 

```{r}

modelstring="
model {

# Likelihood
for (i in 1:nobs) {
lambda[i] <- exp(intercept + alt * altitude[i] + alt2 * altitude[i] * altitude[i] + Ryear[year[i]] + RID[i] ) * Zero[i] + 0.00000001 

beetles[i]~dpois(lambda[i]) 
}

# Fixed effect priors 
intercept ~ dnorm(0,0.0001)
alt ~ dnorm(0,0.0001)
alt2 ~ dnorm(0,0.0001)

# Random effects 

for (i in 1:nyears) {
Ryear[i]~dnorm(0,sigmaYear)
}

for (i in 1:nobs) {
RID[i]~dnorm(0,sigmaID)
}

# Variance priors 
sigmaYear~dgamma(1,2)
sigmaID~dgamma(0.001,0.001)

# Zeroinflation

for (i in 1:nobs) {
Zero[i]~dbern(zeroMu + altZero * altitude[i])
}
zeroMu ~ dunif(0,1)
altZero ~ dnorm(0,0.0001)

# Posterior predictive simulations, conditional on fitted REs  

for (i in 1:nobs) {
  beetlesPredCond[i]~dpois(lambda[i])
}
PredictionCond <- sum(beetlesPredCond)

# Posterior predictive simulations, re-simulating REs  

for (i in 1:nobs) {
  lambdaPred[i] <- exp(intercept + alt * altitude[i] + alt2 * altitude[i] * altitude[i] + RyearPred[year[i]] + RIDPred[i] ) * ZeroPred[i] + 0.00000001 
  beetlesPred[i]~dpois(lambdaPred[i]) 
}

for (i in 1:nyears) {
  RyearPred[i]~dnorm(0,sigmaYear)
}

for (i in 1:nobs) {
  RIDPred[i]~dnorm(0,sigmaID)
}

for (i in 1:nobs) {
  ZeroPred[i]~dbern(zeroMu + altZero * altitude[i])
}

Prediction <- sum(beetlesPred)

}
"

model=jags(model.file = textConnection(modelstring), data=modelData, n.iter=10000,  parameters.to.save = c("intercept", "alt", "alt2", "beetlesPred", "beetlesPredCond", "Ryear", "sigmaYear", "lambda", "altZero", "zeroMu"), DIC = F)

plot(model)

```


```{r}
library(DHARMa)

# Conditional

simulations = model$BUGSoutput$sims.list$beetlesPredCond
pred = apply(model$BUGSoutput$sims.list$lambda, 2, median)
dim(simulations)
sim = createDHARMa(simulatedResponse = t(simulations), observedResponse = data$beetles, fittedPredictedResponse = pred, integerResponse = T)
plotSimulatedResiduals(sim)

# Unconditional

simulations = model$BUGSoutput$sims.list$beetlesPred
pred = apply(model$BUGSoutput$sims.list$lambda, 2, median)
dim(simulations)
# Here I change fittedPredictedResponse because it doesn't make sense to plot the unconditional residuals against the conditional predictions
sim = createDHARMa(simulatedResponse = t(simulations), observedResponse = data$beetles, fittedPredictedResponse = apply(simulations, 2, mean), integerResponse = T)
plotSimulatedResiduals(sim)
```

